% =============================================================
% Mental Model
% Chapter order is set in ../main.tex; the rendered chapter number
% comes from \chapter{} below.
% =============================================================

\chapter{Mental Model}
\label{ch:mental}

\todo{Chapter scaffold. Each section has a one- to three-paragraph
stub indicating intended content. Fleshing out is deferred to a
later writing pass.}

This chapter develops the working mental model that the rest of the
book takes for granted. The angle is deliberately narrow: not how a
transformer is trained, not why attention is a useful operation,
not the historical chain that runs from RNNs to GPT-5. The reader
is assumed to be an engineer who will treat the language model as
a service with a calling convention and who needs to understand
the model's behavior --- its strengths, its limits, the cost
structure of using it --- well enough to design systems that hold
together. We open with the operational picture, work through the
mechanics that follow from it, and close with a taxonomy of
characteristic failures.

\section{What a Language Model Is}
\label{sec:mental-what}

A language model, for the purposes of this book, is a function. The
argument is a sequence of tokens; the value is a probability
distribution over what the next token might be. That is the entire
interface. Everything the engineer does with a model --- chat,
tool calling, retrieval-augmented generation, agentic loops --- is
built on top of this single shape.

\begin{definitionbox}
A \textbf{language model} is a stateless function from a sequence
of tokens to a probability distribution over a fixed vocabulary
of \emph{next} tokens. \emph{Memory} is the text the host hands
the model on the current call; there is no persistent state across
calls. \emph{Generation} is autoregressive: sample a token from
the distribution, append it to the input, repeat.
\end{definitionbox}

\noindent
Two consequences follow from naming the model ``a function'' so
deliberately, and they are worth dwelling on because the rest of
the book leans on both.

The first: a function does not remember. There is no hidden state
carried across calls, no internal variable that knows what the
user said five minutes ago, no ``memory'' in any conventional
sense. The model that processes our second HTTP request is the
same model that processed our first, with no recollection that the
first one happened. Anything the model appears to ``know'' about
the conversation must be in the input it sees on the
\emph{current} call, because there is nowhere else for that
knowledge to live.

The second: a function with a probabilistic output is not, by
itself, a text generator. The model gives us a distribution over
next tokens; turning a distribution into a single next token
requires \emph{sampling}. The runtime samples one token, appends
it to the input, sends the whole thing back to the model, samples
again, repeats. This is autoregressive generation: each output
token feeds back into the input that produces the next one.
Temperature, top-$p$, top-$k$, and the various choices around
them (\S\ref{sec:mental-sampling}) are the knobs on this loop.

\begin{examplebox}
\small
\textbf{One step of the loop.} Given the input
``\code{To be or not to}'' --- a phrase Gemma~4 has seen
verbatim in its training data thousands of times --- the model
returns a distribution sharply peaked on the token \code{ be},
at essentially 100\% probability. Greedy sampling at
temperature zero picks \code{ be}; the host appends it to the
input, producing ``\code{To be or not to be}'', and calls the
model again. The next call's distribution is, perhaps
surprisingly, also dominated by \code{ or} at essentially
100\% --- because the base model continues the parallel
construction it has seen many times in song lyrics, parodies,
and academic paraphrases, not, as one might naively expect,
with ``\code{, that is the question}.'' And so on, token by
token, until the model emits an end-of-sequence token or the
runtime stops it. Why this base-model behavior is \emph{not}
the same as ``the model answers what is being asked'' --- and
what changes when we wrap the input in a chat template --- is
the substance of \S\ref{sec:mental-tokens-practice}.
\end{examplebox}

\noindent
Putting these two facts together: \emph{a language model is a
stateless conditional sampler over a token vocabulary.} That
mouthful is worth internalizing, because it explains every
``memory'' trick the rest of the book will introduce, and every
restriction that follows from one.

The chat history we hold across multi-turn conversations is text
the host appends to the input on every turn. The system prompt
that gives the agent its persona is text at the start of the input
on every turn. The tool advertisements that tell the model about
available functions (Chapter~\ref{ch:toolcalls}) are text in the
input on every turn. The retrieved passages of a RAG system
(Chapter~\ref{ch:context}) are text in the input on every turn.
The model's own previous responses, the user's earlier questions,
the documentation pulled in from a vector store, the reasoning
steps we asked the model to ``think through'' out loud --- all of
it is text in the input on every turn. There is no hierarchy of
importance the model implicitly understands; there is one big
string, lightly segmented by role tags (\S\ref{sec:work-chats}) so
the model can tell which part is which.

Once this picture is in place, a great deal of agent design
becomes legible:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2em]
\item The \emph{context window} of a model is the maximum length
  of this input string. Every byte of system prompt, history,
  tool catalog, and retrieved context counts toward the same
  budget (\S\ref{sec:mental-context}).
\item ``Forgetting'' an old turn is, mechanically, just dropping
  it from the input list before the next call. The model has no
  perspective on what was there before; nothing complains.
\item ``Editing'' the model's own previous response is rewriting
  that part of the input. The model on the next call accepts the
  rewrite as ground truth. Agent harnesses use this to compress
  intermediate results.
\item Agent harnesses with explicit ``memory'' subsystems
  (Chapter~\ref{ch:context}) are deciding, on every turn, what
  to put back in this input.
\end{itemize}

\noindent
This single fact --- the input on this call is the only place
memory lives --- is the most useful piece of mental scaffolding
for everything that follows. We will return to it, sometimes
explicitly, on every page where a design decision is made about
what goes into the model's input.

\subsection{Model types}
\label{sec:mental-types}

Language models differ along several axes that matter for system
design without changing the underlying mental model. The reader
will encounter these dimensions repeatedly throughout the book;
three are consequential enough to name here. We treat the
operational question of which to pick for a given task in
Chapter~\ref{ch:working}; this section just maps the territory.

\paragraph{Multimodal models.}
The interface generalizes naturally beyond text. The input is no
longer a sequence of text tokens but a sequence of \emph{content
blocks}, each typed --- text, image, audio, occasionally video.
The model attends to all of them inside the same context window
and produces text out the other end. By 2026, vision input is
standard across the frontier (OpenAI's GPT family, Anthropic's
Claude, Google's Gemini, Meta's Llama, Google's Gemma); audio
input is standard on most of the same; native image output
remains less common and is more often handled by a separate
specialist model invoked as a tool call
(\S\ref{sec:taxonomy}). The agent-design implication is that
several tools which used to be necessary --- OCR for screenshots,
transcription for voice notes, layout extraction for PDFs --- are
subsumed by a multimodal model directly. The mental model is
unchanged: a stateless conditional sampler over a token
vocabulary, where some of those tokens happen to encode pixels or
audio frames rather than letters.

\paragraph{Open-weights and local models.}
``Open weights'' means the trained parameters of the model are
publicly downloadable, typically under a license that permits
redistribution and fine-tuning. Crucially, ``open weights'' does
\emph{not} mean ``open source'': the training data and training
code are usually not released, and the reader who wants to
reproduce the training run typically cannot. What they
\emph{can} do is download the weights, run them on their own
hardware, fine-tune them on their own data, host them on private
infrastructure, and quantize them for cheaper inference. Each of
those is impossible with a closed-weight hosted model, and each
matters for some class of practical use --- private data,
reproducibility (\S\ref{sec:intro-approach}), regulatory
constraints, latency budgets, cost ceilings. The trade-off is
capability: as of 2026 the strongest open-weights models trail
the strongest closed-weight ones on the hardest benchmarks, but
the gap closes every quarter and for many practical workloads
the open-weights option is competitive. Major open-weights
families to be aware of: Meta's Llama, Alibaba's Qwen, DeepSeek,
Mistral, Microsoft's Phi, Google's Gemma --- the model this book
uses by default. Chapter~\ref{ch:working} treats the choosing-a-model
question in detail; the relevant point here is that ``open
weights'' is a real and increasingly viable design choice, not a
hobbyist concession.

\paragraph{Reasoning / thinking models.}
A relatively recent variant in which the model is trained to
spend extra tokens on a private chain of thought \emph{before}
committing to its final answer. The mechanism is the same
autoregressive loop described above, but the convention is that
some prefix of the output --- often delimited by special tokens,
sometimes hidden from the API consumer --- is ``thinking'' rather
than the user-facing answer; only the post-thinking text is
returned. The trade-off is straightforward. On tasks that benefit
from multi-step reasoning --- mathematics, planning, code that
requires careful tracing, formal verification --- reasoning
models are noticeably stronger, often by a wide margin. On tasks
that don't --- lookup, summarization, simple extraction, format
conversion --- they are merely slower and more expensive, since
the thinking tokens are still tokens the user pays for. Major
families that ship a thinking mode as of 2026 include OpenAI's
o-series, Anthropic's ``extended thinking'', Google's Gemini
``thinking'' variants, DeepSeek-R1, and Qwen-QwQ. We treat
reasoning as one axis in choosing a model
(\S\ref{sec:work-reasoning}), not a separate species; the mental
model of \S\ref{sec:mental-what} carries over unchanged.

\section{Tokens and Tokenization}
\label{sec:mental-tokens}

The model does not see characters and does not see words; it sees
\emph{tokens}. A token is the unit of input the model is trained
on: an integer index into a fixed-size vocabulary, mapped at the
very first layer of the network to a learned vector in the
model's embedding space. Everything downstream --- the attention
pattern, the layer arithmetic, the next-token prediction ---
operates on these vectors. The character ``r'' has no first-class
representation inside the model; its existence is implicit in
whichever tokens happen to contain it, and only to the extent
that the training corpus correlated those tokens with
character-level identity in some downstream task.

\subsection{What to train on}
\label{sec:mental-tokens-what}

The vocabulary cannot grow without bound. Each entry costs an
embedding vector and a row in the output projection, which means
both memory and parameters; the model must be told, before
training begins, that it has exactly $N$ possible tokens to
predict. Two naive choices for what those $N$ tokens should be
present themselves immediately, and both fail in instructive
ways.

\paragraph{Words.}
Tempting, because language is composed of words and a word-level
model would be the most efficient at representing common text:
a sentence of ten words would be a sequence of ten tokens. The
trouble is that the world does not helpfully restrict itself to
a fixed vocabulary. Names, brand identifiers, scientific terms,
neologisms, internet slang, code identifiers, large numbers,
foreign-language fragments, typos --- every one of these
produces ``words'' the model has never seen. A purely word-based
vocabulary either grows unboundedly to chase them all, or falls
back to a single \code{<UNK>} token that collapses every unknown
into the same vector. Neither option is acceptable. Real text
leaks through the cracks of any finite word list, and the model
must have something to say about \emph{strawpberry},
\code{f1bcd2}, \emph{Belfanti}, and \code{quaternion} even when
none of those have appeared in training.

\paragraph{Letters or bytes.}
The opposite extreme. There are only 256 possible byte values,
which fits in a single byte of indexing and admits any input
string without exception. Every hash, every typo, every unknown
name decomposes into bytes the model has seen many times. The
trouble here is the other side of the same coin: the model must
now learn from scratch that \code{t}, \code{h}, \code{e}
together mean \emph{the}; that \emph{the} appears at the start
of countless English sentences; that the bigram \code{th} is
much more common than \code{tg}. The most useful patterns in
language are not at the byte level, they are several bytes long,
and a byte-level model burns much of its capacity reconstructing
those patterns at every position. Sequences also become very
long: a sentence of ten English words might be sixty bytes,
six times the work for the same content.

\subsection{The compromise: bin what's frequent}
\label{sec:mental-tokens-compromise}

The middle ground is to let the data decide. Train the
vocabulary on the same corpus the model will eventually be
trained on; give frequent character sequences their own token,
and reserve single-byte tokens as a fallback for the rare cases.
The resulting vocabulary mixes granularities: \code{the} is one
token because it occurs everywhere; \code{ing} and \code{tion}
are tokens because they recur as suffixes; rare words break into
several pieces; entirely unfamiliar input falls all the way back
to bytes.

The dominant algorithm for producing such a vocabulary is
Byte-Pair Encoding (BPE)~\citep{sennrich2016bpe}, with close
relatives WordPiece and Unigram in production use. Conceptually
the recipe is straightforward.

\begin{enumerate}[label=(\arabic*),leftmargin=2em,itemsep=0.2em]
\item Start with a vocabulary of all single bytes. Every input
  string is representable as a sequence of these.
\item Count, across the training corpus, the frequency of every
  pair of adjacent tokens.
\item Merge the most-common pair into a single new token, added
  to the vocabulary. Re-tokenize the corpus.
\item Repeat until the vocabulary reaches the target size.
\end{enumerate}

\noindent
A run that has seen a great deal of English will discover, very
early, that \code{t} followed by \code{h} is the most common
pair, and merge it into \code{th}; soon after, that \code{th}
followed by \code{e} is the most common pair, and merge into
\code{the}; soon after, that \code{ ing} (with its leading
space) is worth its own token. Run for long enough on a corpus
large enough, and the vocabulary contains common English words
as single tokens, inflectional fragments (\code{ing},
\code{tion}, \code{ly}), most individual letters, and almost
everything in between. Modern vocabularies sit somewhere between
$\sim$50\,000 entries (early GPT-3, Llama~2) and $\sim$256\,000
(Gemma~3 and~4, GPT-4o). At inference time, the tokenizer that
converts an input string into a token sequence is a
longest-match algorithm against this vocabulary.

\subsection{Practical consequences}
\label{sec:mental-tokens-consequences}

Three implications follow from this design and recur throughout
the book.

\begin{itemize}[leftmargin=1.5em,itemsep=0.4em]

\item \textbf{Token counts are not character counts and are not
  word counts.} A useful rule of thumb is that 1000~tokens is
  roughly 750~English words, or about 4000 characters of running
  English prose. The ratio swings widely: code is denser than
  prose because indentation, punctuation, and operators each
  count; structured data (JSON, XML) denser still; rare technical
  vocabulary denser again. A screen of Python is two to three
  thousand tokens; the same screen of plain English is twelve to
  fifteen hundred.

\item \textbf{The cost-per-character is not language-neutral.}
  BPE training corpora have been overwhelmingly English. Latin
  scripts share a tokenizer-friendly structure with English and
  pay a small premium. Cyrillic, Arabic, Devanagari, and CJK
  scripts pay much more: a Mandarin character can be two or three
  tokens, because rare characters fall back to byte-level encoding
  within the tokenizer. On per-token-priced APIs this is a
  multiplicative cost on every interaction in a non-English
  language, and on every prompt that carries non-English text.

\item \textbf{Different models use different tokenizers.} Gemma~4
  ships its own SentencePiece-trained vocabulary of
  $\sim$256\,000 entries; OpenAI's GPT-4 family uses
  \code{cl100k\_base}; GPT-4o uses \code{o200k\_base}; Anthropic
  publishes a \code{count\_tokens} endpoint rather than the
  tokenizer itself. The same string can take 500~tokens on one
  model and 600~on another. Counting accurately --- whether to
  budget context, estimate cost, or fit a passage under a length
  limit --- requires the tokenizer the target model is actually
  using.

\end{itemize}

\noindent
The mechanism is most readable in the small. The example below
uses OpenAI's \code{o200k\_base} (the encoding behind GPT-4o)
because it is freely available; the qualitative pattern is the
same across modern tokenizers.

\begin{examplebox}
\small
\textbf{What the tokenizer actually emits.} Each input below is
followed by its token count and the literal sequence of token
strings that the tokenizer produced.

\begin{itemize}[leftmargin=1.5em,itemsep=0.1em]
\item \code{"hello world"} --- 2 tokens --- \code{["hello", " world"]}
\item \code{"strawberry"} --- 3 tokens --- \code{["st", "raw", "berry"]}
\item \code{"strawpberry"} --- 4 tokens --- \code{["st", "raw", "p", "berry"]}
\item \code{"def count\_letter(word, letter):"} --- 7 tokens --- \code{["def", " count", "\_letter", "(word", ",", " letter", "):"]}
\item \code{"François"} --- 2 tokens --- \code{["Fran", "çois"]}
\item \code{"supercalifragilisticexpialidocious"} --- 11 tokens (rare and broken into syllable-sized pieces)
\item the five-character Mandarin sentence \emph{n\v{i} h\v{a}o,
  sh\`{i}ji\`{e}} (``hello, world'') --- 6 tokens, with the
  comma and the most common characters tokenizing one-to-one and
  the rarer characters falling back to multi-byte sequences
\end{itemize}

\noindent
Note in particular how the misspelling \emph{strawpberry} differs
from the correct \emph{strawberry}: the inserted \texttt{p}
appears as a token of its own because the bigram ``rawp'' is not
in the vocabulary. The model sees the difference between the two
spellings as a four-vector vs three-vector input, not as a
single-character change.
\end{examplebox}

\noindent
The closing observation is the one that motivates the entire
strawberry section of Chapter~\ref{ch:toolcalls}: the model has
no character-level representation of its own input. Any task that
requires reasoning about individual letters --- counting them,
permuting them, checking spelling --- is asking the model to
recover character-level information from token-level vectors,
which it can do only weakly and unreliably. The full unpacking
of this failure mode lives in \S\ref{sec:strawberry}; the short
version here is enough to set up the broader taxonomy of
\S\ref{sec:mental-failures}, where tokenizer-level failures are
one cell in the larger picture.

\section{Tokens and Probabilities in Practice}
\label{sec:mental-tokens-practice}

The previous section described tokens, vocabularies, and BPE in the
abstract. The picture sharpens considerably once one inspects a
specific tokenizer's vocabulary and probes a specific model's
next-token distribution. Both are easy to do offline with an
open-weights model and a local Ollama installation. We use Gemma~4
throughout this section as a representative case; the scripts that
produced the numbers below live in
\fname{manuscript/scripts/tokenization.py} and
\fname{manuscript/scripts/next\_token\_logprobs.py}. The reader is
invited to run them, change the inputs, and watch the distribution
shift in response.

\subsection{Inside Gemma 4's Vocabulary}
\label{sec:mental-tokens-vocab}

Gemma~4 ships a SentencePiece-trained vocabulary of 262{,}145
entries --- 255{,}730 ordinary tokens and 6{,}415 ``added''
tokens reserved for chat-template markers, padding, and similar
protocol bits. The BPE merge list that produced this vocabulary
holds 514{,}906 rules. Inspecting the very first merges is less
illuminating than one might hope, because they are dominated by
long runs of whitespace --- a side-effect of training data dense
with source code and markdown indentation. The first 1{,}393
merges in Gemma~4 are all whitespace-only. After that, the picture
becomes familiar. The first non-whitespace merges are the
highest-frequency English bigrams, in order:
\begin{quote}
  \code{▁t}, \code{in}, \code{▁a}, \code{er}, \code{on},
  \code{he}, \code{re}, \code{en}, \code{at}, \code{▁s},
  \code{or}, \code{▁c}, then \code{▁the}.
\end{quote}

\noindent
This is roughly the canonical English letter-pair frequency
ordering going back to Mayzner and Tresselt. BPE is doing what
it advertises.

Two properties of the resulting vocabulary are worth dwelling
on, because both will reappear in the probability discussion
below.

\paragraph{The longest tokens are sixteen characters.}
The longest entries in Gemma~4's regular vocabulary all cap at
sixteen characters, and fall into three families:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2em]
\item Long English words: \code{circumferential},
  \code{Notwithstanding}, \code{straightforward}.
\item Coding identifiers: \code{InvariantCulture},
  \code{SubDirectoryPath}, \code{FileInputStream}.
\item Spanish and Portuguese words: \code{entretenimiento},
  \code{acontecimientos} (Spanish), \code{disponibilidade}
  (Portuguese).
\end{itemize}

\noindent
That a sixteen-character C\# identifier has its own single
integer ID is, on first encounter, jarring. It is exactly what
the recipe predicts once the input distribution is in view: the
model has been trained on enough .NET code that BPE merged
\code{SubDirectoryPath} all the way up. The identifier is one
token to predict, one token to bill, and one position to attend
over. None of this is the model's intelligence; it is a feature
of the input encoding, frozen at tokenizer-training time.

\paragraph{The leading space lives with the next word.}
SentencePiece marks a leading space with a `\code{▁}' prefix
(U+2581, ``lower one eighth block''). The tokenizer's view of
\code{"Switzerland is Bern"} is
\code{[Switzerland][▁is][▁Bern]} --- each non-initial word's
leading space is part of that word's own token, and there is no
separate inter-word space token in fluent prose. The downstream
consequence is that small variations on a word produce
surprisingly different token sequences. Five concrete examples
make the point:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2em]
\item \code{" Bern"} (with leading space)
  $\to$ \code{[▁Bern]} --- one token. What \code{Bern} looks like
  mid-sentence.
\item \code{"Bern"} (no space)
  $\to$ \code{[Bern]} --- also one token, a separate vocabulary
  entry. The bare form exists for start-of-string and
  post-punctuation contexts.
\item \code{"Bern "} (with trailing space)
  $\to$ \code{[Bern][▁]} --- \emph{two} tokens. Nothing follows
  the trailing space, so it cannot merge into a word-token; it
  becomes a standalone whitespace token.
\item \code{"Berne"} (older French spelling)
  $\to$ \code{[Ber][ne]} --- two tokens. The form is rare enough
  that BPE never gave it a dedicated entry.
\item \code{" Bremen"} $\to$ \code{[▁Bremen]} (one token); but
  \code{"Bremen"} $\to$ \code{[B][remen]} (two). Mid-sentence
  Bremen costs one token; start-of-string Bremen costs two.
\end{itemize}

\noindent
Coverage is uneven even within the small set of European
capitals. \code{" Berlin"} and \code{" Zurich"} are each one
token because training data contains them at saturation;
\code{" Berne"} is two either way. None of this is wrong --- every
string remains representable, falling back to smaller subwords
and ultimately to single bytes --- but the cost in tokens, in
attention, and in the model's prediction problem is not uniform
across what the user would consider ``the same word.''

\subsection{Probabilities in Practice}
\label{sec:mental-tokens-probabilities}

A familiar tutorial cliche goes like this: you type
\code{"The capital of Switzerland is"} into a language model,
and out comes \code{Bern}, because the model has the fact stored
and retrieves it. The cliche is convenient and substantially
misleading. The model is not a fact-retrieval function. It is a
stateless conditional sampler trained to predict the next token
in a stream of text, with no concept of ``question'' until one
has been wrapped around the input by an external chat template.
Three probes against Gemma~4 illustrate what is actually going
on. All numbers below are from the 26B Gemma~4 variant; the
smaller e2b variant produces qualitatively similar shapes with
a flatter distribution.

\paragraph{Probe 1: bare prompt, no chat template.}
Feed the literal string \code{"The capital of Switzerland is"}
to Gemma~4 with the \code{raw=true} flag --- no chat-template
wrapper, no system prompt, exactly the bytes shown. The top four
next-token candidates are:

\begin{center}
\begin{tabular}{lr}
\toprule
token & probability \\
\midrule
\code{" capital"} & 64.5\% \\
\code{" the"}     &  6.3\% \\
\code{" being"}   &  3.8\% \\
\code{" a"}       &  2.9\% \\
\bottomrule
\end{tabular}
\end{center}

\noindent
There is no \code{" Bern"} in the top ten. The model knows about
Bern, perfectly well --- probe~3 will demonstrate it --- but with
no chat template, it has no instruction to give an answer. It is
doing what pre-training trained it to do: predict the next token
in a stream. In actual text, ``The capital of Switzerland is''
is followed by encyclopedic prose: ``the city of Bern, located
in\,\dots,'' ``also the seat of the federal assembly\,\dots,''
``a small city of about 130{,}000 inhabitants\,\dots,'' or ---
driving the top candidate --- ``the capital of the canton of
Bern.'' The token \code{" capital"} reflects exactly that last
pattern, the kind of mildly self-referential sentence (``Bern,
the capital of Switzerland, is the capital of the canton of
Bern.'') that genuinely appears in geography references.

\paragraph{Probe 2: bare prompt plus one trailing space.}
Add a single space --- \code{"The capital of Switzerland is "} ---
and the distribution changes shape entirely:

\begin{center}
\begin{tabular}{lr}
\toprule
token & probability \\
\midrule
\code{"6"}       & 39.5\% \\
\code{"0"}       & 19.5\% \\
\code{"1"}       & 15.2\% \\
\code{"2"}       &  4.5\% \\
\code{"\textbackslash n"} &  4.1\% \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The leading-space property of the tokenizer is the cause. The
trailing space tokenizes as a standalone \code{[▁]} token, and
any candidate that starts with `\code{▁}' would mean two
consecutive spaces in the surface text --- vanishingly rare in
clean prose. So the entire leading-space portion of the
vocabulary --- every word-token --- drops out of the running.
What remains is the no-space portion: digits, punctuation,
control characters. The model picks the contexts in training
data where exactly this configuration appears: enumerated MCQs
(``\code{...is\ \_\_\_\textbackslash n1.\ Bern\textbackslash n2.\ Zurich}\,\dots''),
numeric facts and page numbers (``\code{...is\ 134{,}591\,\dots}''),
and scraped text where a trailing space was an editorial slip
before a line break. The user who typed the trailing space
``meant'' nothing by it; the model has no choice but to show
what the byte sequence actually leads to in the wild.

\paragraph{Probe 3: chat template, thinking suppressed.}
Wrap the same question in Gemma~4's chat template. What the model
actually receives is:

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize,numbers=none,frame=single,framerule=0.4pt,xleftmargin=1em,xrightmargin=1em]
<start_of_turn>user
What is the capital of Switzerland? Answer in one word.<end_of_turn>
<start_of_turn>model
\end{lstlisting}

\noindent
Disable the thinking channel via Ollama's \code{think: false}
flag on its native \code{/api/chat} endpoint (the
OpenAI-compatible \code{/v1/chat/completions} silently ignores
the flag). The first-token distribution at the answer position is:

\begin{center}
\begin{tabular}{lr}
\toprule
token & probability \\
\midrule
\code{"Bern"}    & 100.00\% \\
\code{"Ber"}     &   0.00\% \\
\code{" Bern"}   &   0.00\% \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Bern at one hundred percent. The model knew the answer the
entire time. The difference between probes~1 and~3 is not
knowledge but input scaffolding: a chat template wrapping the
question, and a flag that suppresses the model's thinking
channel so the first token is the answer rather than a
reasoning preamble.

\subsubsection*{What models are actually trained on}

The gap between probes 1 and 3 traces to the layered structure
of modern LLM training. Four stages, roughly:

\begin{enumerate}[leftmargin=1.5em,itemsep=0.2em]
\item \textbf{Pre-training} on a multi-trillion-token corpus of
  web text, books, and source code, with the next-token-prediction
  objective. This produces the base model. Probes 1 and~2 are
  reading directly off this layer.
\item \textbf{Supervised fine-tuning (SFT)} on hundreds of
  thousands to millions of human-curated dialog examples. User
  turns are wrapped in role-tagged chat templates; assistant
  responses follow a particular style. The chat-template tokens
  --- \code{<start\_of\_turn>}, \code{<end\_of\_turn>}, the role
  markers --- appear here for the first time as meaningful
  structure rather than as random byte sequences.
\item \textbf{Reinforcement learning from human feedback (RLHF)},
  or its AI-rater cousin RLAIF. The model is shown competing
  candidate responses, picks one, and is rewarded or penalized
  according to rater preference. This tunes tone, helpfulness,
  and refusal behavior.
\item \textbf{Reasoning specialization.} A more recent training
  pass that introduces a private ``thinking'' channel --- marked
  with its own structural tokens --- in which the model writes
  intermediate reasoning before producing a user-facing answer.
  Suppressing it is what the \code{think: false} flag in probe~3
  is for.
\end{enumerate}

\noindent
The control sequences that distinguish probe 3 from probe 1 ---
\code{<start\_of\_turn>}, \code{<end\_of\_turn>}, the
thinking-channel markers --- are artifacts of stages~2--4. The
base model has never been instructed that ``this is a question,
answer it.'' What it has done is read enormous amounts of text.
The post-training stages teach it: ``when wrapped in this
template, your job is to produce an answer in this style.''
Without the wrapper, the prior remains the pre-training prior:
predict the next token in a continuation. The wrapper is
something the host application controls --- not the model.

\subsubsection*{Three takeaways}

\begin{itemize}[leftmargin=1.5em,itemsep=0.3em]

\item \textbf{Token probabilities are conditional distributions,
  not facts.} They are conditioned on a specific token sequence,
  sampled from a specific training distribution. They report the
  model's continuation prior, not its beliefs.

\item \textbf{One byte of input can swing the output.} Probes 1
  and 2 differ by a single trailing space and produce radically
  different distributions. Tokenizer mechanics propagate.
  Tutorials that wave at tokenization as ``a detail you don't
  need to worry about'' are misleading you.

\item \textbf{Chat templates and thinking channels are part of
  the system, not the model.} The reason ``The capital of
  Switzerland is'' does not produce ``Bern'' while a chat-wrapped
  question does is not a model deficit but a missing wrapper.
  The engineer extracting reliable behavior from an LLM must
  reason at the granularity of ``what token sequence is the model
  seeing right now,'' not ``what question is the user asking.''
  The two are related, but they are not the same.

\end{itemize}

\begin{asidebox}[title={Aside --- Choosing the ``best'' sequence}]
\small
The probes above quietly assume that the most-probable next token
is the ``right'' one --- and, by extension, that the most-probable
\emph{sequence} would be the best output. The actual story is less
tidy. A line of work in the neural-translation literature, summarised
under the name \emph{inadequacy of the mode}~\citep{eikema2020inadequacy},
showed that for the majority of test sentences the human
gold-reference translation receives \emph{lower} model probability
than the model's own most-probable output. The global maximum of a
trained sequence model's distribution is often not where the best
text lives. We will treat sampling-based decoding in
\S\ref{sec:mental-sampling} without belabouring the point; the
interested reader is warmly invited to follow the reference.
\end{asidebox}

\section{The Context Window}
\label{sec:mental-context}

\todo{Short. Nominal context vs effective context: the
``lost-in-the-middle'' phenomenon, the empirical observation that
recall on facts placed in the middle of a long context degrades
sharply, and that effective utility caps out well below the
advertised maximum. Forward-reference the
Tool-Calls-and-Context section of Chapter~\ref{ch:toolcalls} for
the agent-side implications. Detail of \emph{what fills} the
context (system prompt, tool catalog, history, tool results)
lives there, not here.}

\section{Sampling and Determinism}
\label{sec:mental-sampling}

\todo{What temperature, top-k, top-p are at the level of intuition.
Greedy decoding at temperature zero. The catch: ``temperature 0''
is not actually deterministic across providers, because of
batching, hardware variation, and silent model updates. This has
direct consequences for testing agents
(Chapter~\ref{ch:reliable}). Mention seed parameters where they
exist and where they do not.}

\section{Embeddings}
\label{sec:mental-embeddings}

\todo{Short. An embedding model maps a piece of text to a
fixed-length vector such that semantically similar texts get
nearby vectors. Practical facts the reader needs: typical
dimensionalities (512, 768, 1024, 1536); cosine similarity / dot
product as the comparison function; dramatically cheaper per call
than chat models; the existence of dedicated embedding models
distinct from chat models. Defer the \emph{use} of embeddings
(vector stores, retrieval, memory, deduplication) to the
agent-context chapter where RAG is treated. Frame here as
``another thing the same family of models can do, with a different
output shape.''}

\section{Strengths and Failure Modes}
\label{sec:mental-failures}

\todo{The substantive section. A small taxonomy of what language
models are good at and what they consistently fail at, with the
mechanism for each failure named. This is the section that pays
for the rest of the chapter --- the engineer who has internalized
it will design systems differently.}

\subsection{What models are good at}
\label{sec:mental-strengths}

\todo{One paragraph. Strong: language production and comprehension;
code; pattern completion; well-known facts that were dense in
training data; transformations and summarizations of given text;
in-context learning from a few examples.}

\subsection{Tokenizer-level failures}
\label{sec:mental-tok-fail}

\todo{Counting characters, manipulating individual letters, exact
arithmetic on long numbers. Forward-reference
\S\ref{sec:strawberry}, which works the canonical example through.}

\subsection{Hallucination}
\label{sec:mental-hallucination}

\todo{The model produces plausible content that is not true. Why
this happens (the training objective rewards plausibility, not
truth). Mitigations: tools that ground the model in real data
(\S\ref{sec:taxonomy}), retrieval, verifiers (the Lean example
from \S\ref{sec:intro-now}).}

\subsection{Sycophancy and instruction drift}
\label{sec:mental-sycophancy}

\todo{The model agrees too readily with the user's framing, even
when the framing is wrong. Related: over long contexts the model
drifts away from the original instruction in the system prompt.
Implications for agent design: do not assume that pushback
survives disagreement; periodically re-anchor the system prompt.}

\subsection{Sensitivity to phrasing and formatting}
\label{sec:mental-sensitivity}

\todo{Small changes to the prompt --- punctuation, ordering of
examples, capitalization --- can produce non-trivially different
outputs. This makes evaluation subtle (Chapter~\ref{ch:reliable})
and prompts brittle. Forward-reference the prompting-craft
chapter (Chapter~\ref{ch:tools}).}

\subsection{Distributional and adversarial fragility}
\label{sec:mental-fragility}

\todo{Models perform well on inputs that resemble their training
distribution and worse on inputs that do not. Adversarial
inputs --- including prompt injection, treated in
Chapter~\ref{ch:reliable} --- exploit this gap.}

% Structured Outputs has been moved to Chapter~\ref{ch:working},
% \S\ref{sec:structured}. It is more naturally a property of how
% one calls a model than of the model itself.
