% =============================================================
% Working with LLMs
% Chapter order is set in ../main.tex; the rendered chapter number
% comes from \chapter{} below.
% =============================================================

\chapter{Working with LLMs}
\label{ch:working}

\todo{Chapter scaffold. The setup-and-wrapper sections are written
in full; the surrounding sections are stubs. Where the material is
operational and dates fast --- model lineups, prices, SDK names ---
the stub also notes how to write it so the chapter ages gracefully.}

Where Chapter~\ref{ch:mental} addressed \emph{what a language model
is}, this chapter addresses \emph{how to call one}. The treatment
is operational. We name the players, give the criteria for choosing
among them, set the reader up to actually run a model in one of
three deployment shapes, define the \code{query\_llm} wrapper that
every later script in the book will import, walk through the
chat-message structure that every modern API exposes, treat
structured outputs as the specialization of message structure that
powers tool calls, and close with the cost, caching, and latency
considerations that shape every production system. None of this is
theoretically deep; all of it is practically necessary.

\section{Model Providers}
\label{sec:work-providers}

\todo{Short, deliberately. Two parts.}

\paragraph{The landscape.}
\todo{As of 2026, the hosted frontier is dominated by three
providers --- OpenAI (GPT family), Anthropic (Claude family),
Google (Gemini family). The open-weights side has its own gravity:
Meta's Llama, Alibaba's Qwen, DeepSeek, Mistral, Microsoft's Phi,
and Google's Gemma straddling both camps. The point of this
paragraph is to name the players the reader will see referenced
elsewhere in the book; it is not to rank them.}

\paragraph{Axes that age slowly.}
\todo{Rather than a player-by-player capability table, list the
axes on which providers differentiate themselves: hosted vs
open-weights; frontier capability vs commodity-cheap; generalist
vs reasoning-specialized; long context vs strong-but-short;
multimodal vs text-only; permissive vs strict on content. Axes
age much more slowly than rankings, and the reader will use them
to think about whichever providers exist when they are reading.}

\section{Choosing a Model}
\label{sec:work-choose}

\todo{The decision criteria. Capability needed for the task;
context budget; latency tolerance; cost ceiling; whether the data
must stay on premises; whether thinking-mode helps. The
recommended default: use a small, fast, cheap model where you can
and route to a strong one only when you must. Mention orchestration
patterns briefly --- a router model that decides which strong model
sees the task --- and forward-reference the agents chapters where
multi-model orchestration is treated.}

\subsection{Reasoning models in the choice}
\label{sec:work-reasoning}

\todo{Treat ``reasoning'' as one axis of the choice rather than a
separate section. When extended thinking helps (multi-step
mathematics, planning, code that requires careful tracing); when
it does not (lookup, summarization, simple extraction). Cost
trade-off and latency implications.}

\section{Chats: the Message Structure}
\label{sec:work-chats}

The HTTP API every modern provider exposes is, on the surface,
a \emph{chat} endpoint. A request carries not one prompt but a
\emph{list of messages}, each tagged with a role. A response is a
new message, also tagged with a role, that the runtime is
expected to append to the list before the next request. This
structure has the look of a database schema and the importance of
a calling convention: it determines what we can ask the model
and what it can answer with. Before writing any code that calls a
model, we need to know what is in this list and why.

This section is purely about the shape. Prompting craft --- what
to actually \emph{say} in the messages --- belongs to
Chapter~\ref{ch:tools}.

\subsection{Why a list, and not a string}
\label{sec:work-chats-why}

Chapter~\ref{ch:mental} established that a language model is a
stateless function from a token sequence to a probability
distribution over next tokens. It has no memory; everything it
``knows'' about the conversation must be in the input tokens of
the current request. The chat-message list is the convention by
which the runtime communicates the structure of that input: which
parts came from the developer, which parts came from the user,
which parts the model itself produced earlier, which parts are
results from tools the model called.

Under the hood, the list is collapsed back into a single token
sequence using a \emph{chat template} specific to the model.
Gemma~4 inserts the special tokens \code{<\textbar turn>} and
\code{<turn\textbar >} around each role boundary; Anthropic's
hosted models use XML-shaped delimiters; OpenAI keeps the
template internal and never exposes it. The host that calls the
API does not need to know any of this, which is the point of the
abstraction: the list of role-tagged dicts is the contract, and
the model serializes it however it must.

\subsection{The four roles}
\label{sec:work-chats-roles}

Every modern provider supports three roles at minimum, plus a
fourth introduced once tool calls enter the picture.

\begin{description}[style=nextline,leftmargin=1.5em,itemsep=0.4em]

\item[\code{system}] Instructions \emph{from the author of the
  agent} that frame the entire conversation. Persona, tone,
  format requirements, the catalog of tools the model is
  allowed to call (Chapter~\ref{ch:toolcalls}), policy
  constraints. The system message is conceptually the model's
  ``standing orders'' for this session. There is typically
  exactly one, at the start of the list.

\item[\code{user}] What the human (or, in an agent context,
  whichever upstream system is acting as the user) is asking
  for in the current turn. Questions, instructions, pasted
  content, retrieved context.

\item[\code{assistant}] What the model itself has produced
  earlier in this conversation. The runtime appends an
  \code{assistant} entry after every model response; on the
  next request, that entry is back in the input as part of the
  history.

\item[\code{tool}] The result of a tool call. Introduced in
  Chapter~\ref{ch:toolcalls}, where we work through the
  tool-call exchange in detail. Until then, the three roles
  above are sufficient.

\end{description}

\noindent
A first-turn conversation is therefore the smallest non-trivial
list:

\begin{lstlisting}[language=json,caption={The smallest useful messages list: one system frame, one user question.},label=lst:chat-first-turn]
[
  {"role": "system", "content": "You are a helpful assistant that answers in one short sentence."},
  {"role": "user",   "content": "What is the capital of Switzerland?"}
]
\end{lstlisting}

\noindent
The model's response (in the OpenAI shape) is one
\code{assistant} message:

\begin{lstlisting}[language=json,caption={The response from the model: a single assistant message.},label=lst:chat-first-response]
{
  "role":    "assistant",
  "content": "The capital of Switzerland is Bern."
}
\end{lstlisting}

\noindent
Provider-specific response wrappers vary --- OpenAI nests this
inside \code{choices[0].message}; Anthropic returns a list of
content blocks under \code{content}; Ollama uses
\code{message} directly --- but the inner shape is the same
\code{\{role, content\}} dict.

\subsection{Multi-turn conversations}
\label{sec:work-multiturn}

A multi-turn conversation is an append-only list. After each
exchange the runtime appends two entries: one \code{user} (the
new question) and one \code{assistant} (the previous model
response that prompted it). On the next request, the entire list
is re-sent to the model. This is the operational consequence of
the model being stateless across HTTP requests: the only place
``conversation history'' lives is inside the messages list that
the host hands the model.

\begin{lstlisting}[language=json,caption={A three-turn conversation, immediately before the model's reply to the third user message is generated.},label=lst:chat-multiturn]
[
  {"role": "system",    "content": "You are a helpful assistant ..."},
  {"role": "user",      "content": "What is the capital of Switzerland?"},
  {"role": "assistant", "content": "The capital of Switzerland is Bern."},
  {"role": "user",      "content": "And how many people live there?"},
  {"role": "assistant", "content": "Bern has about 145,000 inhabitants."},
  {"role": "user",      "content": "What about the metro area?"}
]
\end{lstlisting}

\noindent
Two consequences of this design that the practitioner should
internalize.

First, \emph{cost grows with conversation length}. Every turn
re-sends the entire history, so the input-token bill on turn~$n$
includes the cost of all $n-1$ previous turns. Long conversations
become expensive in a way that linearly increases per turn.
\S\ref{sec:work-cost} treats prompt caching, which is the
provider-side response to this growth.

Second, \emph{the runtime is in charge of memory}. ``Forgetting''
an earlier turn is just dropping it from the list; ``editing''
the model's own previous response is just rewriting the
\code{assistant} entry in place. The model has no opinion about
either. Agent harnesses with explicit memory management
(Chapter~\ref{ch:context}) lean heavily on this freedom.

\subsection{The system prompt in practice}
\label{sec:work-system}

The system prompt is where most of the agent's character lives. A
useful working rule:

\begin{quote}
\itshape
Stable instructions go in the system message; per-turn content
goes in the user message.
\end{quote}

\noindent
``Stable'' here means \emph{constant across the entire
session}. The model's persona, the response format, the list of
tools it may call, the policy constraints under which it
operates --- these belong in \code{system}. The actual question
the user is asking right now belongs in \code{user}.

The distinction matters for two operational reasons. The first is
caching: most providers offer prompt caching that hits when the
prefix of the messages list is identical across requests
(\S\ref{sec:work-cost-cache}). A stable system prompt followed by
a variable user message is exactly the layout that lets prompt
caching pay off. Mixing per-turn content into \code{system}
breaks the cache.

The second is interpretation: many providers train their models
to weight \code{system} instructions more strongly than
\code{user} instructions, on the grounds that the developer is
more trusted than the end user. The defense-in-depth pattern of
Chapter~\ref{ch:reliable} relies on this asymmetry: the
\code{system} prompt sets the rules; an end user (or worse, an
adversarial input retrieved from the web) cannot easily override
them just by saying ``ignore previous instructions''.

The deeper craft of writing system prompts that work --- few-shot
examples, role definitions, what to put first, what to put last
--- is the subject of Chapter~\ref{ch:tools}, and we leave it
there.

\section{Interacting with LLMs}
\label{sec:work-deploy}

With the message structure in hand from \S\ref{sec:work-chats}, we
can now look at the calling mechanics. Two questions remain:
\emph{where does the model run?} and \emph{how does the call go
out and come back?} The three subsections that follow this
introduction work the first question through in detail, one
deployment shape at a time. The second question is treated here
at the level of orientation; the \code{query\_llm} wrapper of
\S\ref{sec:work-wrapper} is the working answer that the rest of
the book imports.

\paragraph{Where the model runs.}
There are three deployment shapes that a practical reader will
encounter. The choice between them is mostly orthogonal to which
model is being served; a reader who has any one of the three
working can run any of the book's later code.

\begin{description}[style=nextline,leftmargin=1.5em,itemsep=0.3em]
\item[Hosted API.] The most common option, and the simplest to
  start with. OpenAI, Anthropic, and Google expose chat-completion
  APIs over HTTPS; we never see where the model actually runs, we
  just hit an endpoint with an API key.
\item[Local model server.] The model runs on the reader's own
  hardware. We still call it via an HTTP API --- typically Ollama,
  which serves a chat-completion endpoint indistinguishable in
  shape from the hosted ones --- but the server is a process on
  the local machine.
\item[Remote model server.] Identical to the local pattern, except
  the GPU lives in someone else's data center. The reader rents a
  GPU instance pre-loaded with Ollama, vLLM, or
  text-generation-inference; the resulting URL is what the
  \code{query\_llm} wrapper of \S\ref{sec:work-wrapper} is then
  pointed at. From the calling code's perspective the only
  difference from the local case is the value of one environment
  variable.
\end{description}

\subsection{The anatomy of an LLM call}
\label{sec:work-anatomy}

Across all three deployment shapes, the calling pattern is the
same: we POST a JSON body to an HTTP endpoint, and we read a JSON
body back. The function we are building toward, which the rest of
the book will import, has a signature of roughly the following
shape:

\begin{lstlisting}[language=Python,caption={The shape of the wrapper we are heading for. The full implementation is in \S\ref{sec:work-wrapper}.},label=lst:query-llm-simple]
def query_llm(messages,
              model:       Optional[str] = None,
              temperature: float = 0.0,
              max_tokens:  int = 2048) -> str:
    """Send `messages` to the configured LLM and return its reply."""
    ...
\end{lstlisting}

\noindent
This subsection looks at the two halves of that call --- what we
send out, and what we get back --- in turn. The smoke tests of
\S\S\ref{sec:work-deploy-hosted}--\ref{sec:work-deploy-remote}
will then exercise both halves against real endpoints.

\paragraph{Request.}
Three of the inputs are mandatory across every provider, and two
are optional but almost always supplied.

\begin{description}[style=nextline,leftmargin=1.5em,itemsep=0.3em]
\item[\code{messages}] The list of role-tagged dicts of
  \S\ref{sec:work-chats}. This is what the function carries
  through to the underlying API; nothing else in the call shape
  depends on whose backend is in use.
\item[\code{model}] A string identifier picking which model to
  use. On hosted APIs the choice is often between several
  frontier models and several cheaper variants; on a local Ollama
  the choice is whatever has been pulled with \code{ollama pull}.
\item[An API key.] For hosted backends, the credential that says
  who is paying for these tokens. For a local Ollama there is no
  authentication; for a remote server it depends on how the
  provider has set up access control. The wrapper reads the key
  out of an environment variable so that scripts are never
  committed with credentials embedded.
\item[\code{temperature}] The sampling parameter introduced in
  \S\ref{sec:mental-sampling}. Zero gives the most-determined
  output; higher values trade reproducibility for variety. We
  default to zero throughout the book.
\item[\code{max\_tokens}] An upper bound on generated tokens. Two
  reasons it matters: it caps cost on hosted APIs, and it caps
  latency in agentic loops where a runaway generation would
  otherwise block the next iteration.
\end{description}

\paragraph{Response.}
What comes back is a JSON object. It contains rather more than the
assistant's reply, and it pays to know what each piece is for
before reaching past it.

The most important piece, naturally, is the \emph{assistant's
message itself} --- a \code{\{"role": "assistant", "content":
"\dots"\}} dict in exactly the shape we saw in
\S\ref{sec:work-chats-roles}. The runtime is expected to take this
dict and append it to the messages list before the next request,
which is how multi-turn conversations grow
(\S\ref{sec:work-multiturn}).

Wrapped around the message, the response carries several
operational fields the practitioner will use sooner or later:

\begin{description}[style=nextline,leftmargin=1.5em,itemsep=0.3em]
\item[Token counts.] How many input tokens were consumed and how
  many output tokens were generated. These are what billing is
  computed against on hosted backends (\S\ref{sec:work-cost-cost})
  and what one tracks when monitoring an agent's context budget
  in long-running sessions.
\item[A finish reason.] Why generation stopped. The common values
  are \code{stop} (the model produced an end-of-sequence token of
  its own accord), \code{length} (it hit the \code{max\_tokens}
  cap and was cut off mid-sentence), and, in
  Chapter~\ref{ch:toolcalls}, \code{tool\_calls} (it produced a
  structured tool-call request rather than text). Distinguishing
  these is essential for debugging: a truncated reply is not the
  same kind of failure as a refusal.
\item[A response identifier and timing block.] Provider-specific.
  Useful for logs (correlating a response with a request in a
  trace) and, on backends that support it, for re-fetching or
  resuming a generation.
\end{description}

\noindent
On top of all this, OpenAI-flavored responses add one more layer
of wrapping that confuses the reader the first time it shows up:
the \emph{choices} field. The chat-completion endpoint can be
asked to generate $n > 1$ candidate completions for the same
input --- each an independent sample, drawn at the same
temperature --- and each candidate is one entry in a
\code{choices} array, with its own \code{message} and its own
\code{finish\_reason}. In the common case of $n = 1$ (which is
the default and which the book uses throughout),
\code{choices[0]} is the only entry and it carries the assistant
message we want.

This is why the smoke tests in
\S\S\ref{sec:work-deploy-hosted}--\ref{sec:work-deploy-remote}
end with the drill-down

\begin{lstlisting}[language=Python]
print(rsp.json()["choices"][0]["message"]["content"])
\end{lstlisting}

\noindent
Reading right to left: \code{rsp.json()} parses the JSON
response; \code{["choices"][0]} picks the first (only) candidate;
\code{["message"]} is the role/content dict; \code{["content"]}
is the assistant's text. It is the standard idiom for ``give me
the model's reply as a string,'' and one of the things
\code{query\_llm} of \S\ref{sec:work-wrapper} will hide.

The other backends look slightly different at the top of the
wrapper but ultimately reach the same place: Anthropic returns
content blocks under \code{content}
(\code{rsp.json()["content"][0]["text"]}); Ollama uses
\code{message} directly without the \code{choices} array
(\code{rsp.json()["message"]["content"]}). Each is one
provider-specific path to the same $\{$role, content$\}$ dict.

For the smoke tests below, and for the rest of the book's simple
tests, this drill-down is all we will use. Token counts, finish
reasons, and identifiers are the next layer up; we surface them
in Chapter~\ref{ch:reliable} where production observability
demands it. Until then, the assistant's text is what we want, and
\code{["choices"][0]["message"]["content"]} is how we get it.

\subsection{The hosted API}
\label{sec:work-deploy-hosted}

The simplest route is to pay one of the major providers per token
and POST to their HTTPS endpoint. The provider's API documentation
gives the specifics --- the URL, the header conventions, the
exact field names --- and they all look broadly the same.

\begin{lstlisting}[language=bash,caption={Hosted-API setup, OpenAI flavor.},label=lst:setup-hosted]
$ pip install requests
$ export OPENAI_API_KEY=sk-...
\end{lstlisting}

A minimal smoke test:

\begin{lstlisting}[language=Python,caption={Smoke test against the OpenAI chat-completion endpoint.},label=lst:smoke-hosted]
import os, requests

rsp = requests.post(
    "https://api.openai.com/v1/chat/completions",
    headers={"Authorization": f"Bearer {os.environ['OPENAI_API_KEY']}"},
    json={
        "model":    "gpt-4o-mini",        # any current chat model
        "messages": [{"role": "user", "content": "say hi in three words"}],
    },
    timeout=60,
)
print(rsp.json()["choices"][0]["message"]["content"])
\end{lstlisting}

\noindent
If this prints something like \emph{``Hi there, friend!''}, the
hosted route is working. The Anthropic and Gemini equivalents
differ only in the URL, header conventions, and one or two field
names; we treat all three uniformly in \S\ref{sec:work-wrapper}.

When this is the right choice: you do not have a GPU, your task
needs a frontier capability, you do not require pinned weights, the
budget is acceptable. The disadvantages were treated in
\S\ref{sec:intro-approach}: hosted models change underneath their
identifiers, and reproducibility against a fixed identifier is not
guaranteed.

\subsection{Local with Ollama}
\label{sec:work-deploy-local}

Run a model on your own machine. The book's default working
configuration is Gemma~4 served via Ollama, for the reproducibility
reasons of \S\ref{sec:intro-approach}. Setup is two steps:

\begin{enumerate}[label=(\arabic*)]
\item Install Ollama. Download from \url{https://ollama.com/download}
  (single binary on macOS, Linux, Windows; or \code{brew install
  ollama} on macOS).
\item Pull a model:

\begin{lstlisting}[language=bash]
$ ollama pull gemma4:e4b      # small, fits on a laptop
$ ollama pull gemma4:26b      # large, needs a real GPU
\end{lstlisting}
\end{enumerate}

\noindent
Ollama listens on \url{http://localhost:11434} by default. A smoke
test:

\begin{lstlisting}[language=Python,caption={Smoke test against a local Ollama server.},label=lst:smoke-local]
import requests

rsp = requests.post(
    "http://localhost:11434/api/chat",
    json={
        "model":    "gemma4:e4b",
        "stream":   False,
        "messages": [{"role": "user", "content": "say hi in three words"}],
    },
    timeout=120,
)
print(rsp.json()["message"]["content"])
\end{lstlisting}

\noindent
When this is the right choice: you have a GPU or Apple Silicon
machine with enough memory, you want pinned-weights
reproducibility, your data is sensitive enough that an external API
is a non-starter, or you prefer a flat hardware cost over metered
per-token billing. The disadvantages: smaller models are noticeably
less capable on the hardest tasks, and the larger Gemma sizes will
not fit on consumer GPUs.

A rough hardware sanity guide for Gemma~4:

\begin{description}[style=nextline,leftmargin=2em,itemsep=0.2em]
\item[\textnormal{Apple Silicon, 16~GB unified memory:}]
  \code{gemma4:e2b} comfortable, \code{gemma4:e4b} usable, anything
  larger swaps.
\item[\textnormal{Apple Silicon, 32~GB unified memory:}]
  \code{gemma4:e4b} comfortable, \code{gemma4:26b} runs but is slow.
\item[\textnormal{NVIDIA GPU, 8~GB VRAM:}] \code{gemma4:e2b} only.
\item[\textnormal{NVIDIA GPU, 24~GB VRAM:}] \code{gemma4:e4b}
  comfortable, \code{gemma4:26b} usable.
\item[\textnormal{NVIDIA GPU, 80~GB VRAM (data-center class):}]
  everything, including reasoning-mode variants.
\end{description}

\noindent
Numbers are approximate and quantization-dependent. The reader who
exceeds their hardware by one tier will get correct outputs at
unreasonable speed; the reader who exceeds by two tiers will run
out of memory.

\subsection{Remote model server}
\label{sec:work-deploy-remote}

The third deployment shape is the local model server's pattern,
except the GPU lives in someone else's data center. You rent a GPU
instance with Ollama (or vLLM, or text-generation-inference)
pre-installed; you get back a URL; the rest of the book's code,
including the \code{query\_llm} wrapper of
\S\ref{sec:work-wrapper}, points at that URL and otherwise behaves
identically to the local case.

This is the right choice when you want pinned open weights but
cannot fit the model on your own hardware. RunPod and Vast charge
by the hour at rates that, for an idle box, are higher than per-token
hosted APIs but for a workload with steady throughput are
competitive. Together and Replicate are slightly different: they
expose hosted open-weight models via APIs (closer in shape to
OpenAI than to Ollama), but the model versions are pinned and the
weights are downloadable elsewhere.

A typical setup with RunPod:

\begin{enumerate}[label=(\arabic*)]
\item Pick a GPU template that has Ollama pre-installed (or use the
  bare Ubuntu template and install it manually).
\item Expose port 11434.
\item Pull the model on first boot
  (\code{ollama pull gemma4:26b}).
\item Point the book's \code{OLLAMA\_BASE\_URL} environment
  variable at the RunPod-issued URL.
\end{enumerate}

\noindent
The smoke test is identical to Listing~\ref{lst:smoke-local} with
\code{localhost:11434} replaced by the public URL. The
\code{query\_llm} wrapper handles the URL substitution
transparently.

\section{Your First Call: a query\_llm Wrapper}
\label{sec:work-wrapper}

The three deployment shapes of \S\ref{sec:work-deploy} expose
subtly different HTTP endpoints. Rather than scatter the
differences across every script in the book, we define a single
\code{query\_llm} function that hides them. Every later chapter
imports this function; the reader can switch between hosted and
local backends with a one-line change in a \fname{.env} file.

The wrapper sits at \fname{manuscript/scripts/query\_llm.py}; the
salient pieces are inline below. It is deliberately minimal. There
is no retry logic, no exponential back-off, no token counting, no
streaming, and no tool-call handling. Each of those would be
useful in a production wrapper; none of them belongs in the
function the book uses for thirty pages of teaching. Tool calls
are introduced in Chapter~\ref{ch:toolcalls}, which builds a
richer wrapper on the same backend functions defined here.

\paragraph{Configuration.}
The wrapper reads its configuration --- which backend, which
default model, which API key --- from a \fname{.env} file in the
script's directory, via the \code{python-dotenv} package. A
template file \fname{.env.example} ships alongside; copying it to
\fname{.env} and editing it is the entire setup:

\begin{lstlisting}[language=bash,caption={Setting up \code{query\_llm}.},label=lst:wrapper-setup]
$ pip install requests python-dotenv
$ cp .env.example .env
$ $EDITOR .env       # uncomment and fill the lines you need
\end{lstlisting}

\noindent
Keeping credentials in a file rather than asking the reader to
\code{export} them on every shell session has two practical
benefits beyond convenience: shell history does not log them, and
the file can be \code{git-ignore}d in one place
(\fname{ai\_vault/.gitignore} already does this). Shell-set
environment variables, where present, take precedence over the
\fname{.env} file --- useful for one-off overrides.

\paragraph{The wrapper.}

\begin{lstlisting}[language=Python,caption={\fname{query\_llm.py} --- a three-backend chat wrapper.},label=lst:query-llm]
import os
from typing import Optional

import requests
from dotenv import load_dotenv

load_dotenv()   # reads .env into os.environ at import time

def query_llm(
    messages: list[dict[str, str]],
    *,
    model:       Optional[str] = None,
    temperature: float = 0.0,
    max_tokens:  int = 2048,
) -> str:
    """Send `messages` to whichever backend LLM_BACKEND selects.

    LLM_BACKEND is one of {"ollama", "openai", "anthropic"};
    defaults to "ollama". Returns the assistant's text response.
    """
    backend = os.environ.get("LLM_BACKEND", "ollama")
    dispatch = {
        "ollama":    _ollama_chat,
        "openai":    _openai_chat,
        "anthropic": _anthropic_chat,
    }
    if backend not in dispatch:
        raise ValueError(f"unknown LLM_BACKEND: {backend!r}")
    return dispatch[backend](messages, model, temperature, max_tokens)


def _ollama_chat(messages, model, temperature, max_tokens) -> str:
    base = os.environ.get("OLLAMA_BASE_URL", "http://localhost:11434")
    rsp = requests.post(
        f"{base}/api/chat",
        json={
            "model":    model or os.environ.get("OLLAMA_MODEL", "gemma4:e4b"),
            "stream":   False,
            "messages": messages,
            "options":  {"temperature": temperature, "num_predict": max_tokens},
        },
        timeout=120,
    )
    rsp.raise_for_status()
    return rsp.json()["message"]["content"]


def _openai_chat(messages, model, temperature, max_tokens) -> str:
    rsp = requests.post(
        "https://api.openai.com/v1/chat/completions",
        headers={"Authorization": f"Bearer {os.environ['OPENAI_API_KEY']}"},
        json={
            "model":       model or "gpt-4o-mini",
            "messages":    messages,
            "temperature": temperature,
            "max_tokens":  max_tokens,
        },
        timeout=120,
    )
    rsp.raise_for_status()
    return rsp.json()["choices"][0]["message"]["content"]


def _anthropic_chat(messages, model, temperature, max_tokens) -> str:
    # Anthropic puts the system message in a separate field rather
    # than in the messages list, so we split it out.
    system = next((m["content"] for m in messages if m["role"] == "system"), "")
    msgs   = [m for m in messages if m["role"] != "system"]
    rsp = requests.post(
        "https://api.anthropic.com/v1/messages",
        headers={
            "x-api-key":         os.environ["ANTHROPIC_API_KEY"],
            "anthropic-version": "2023-06-01",
            "content-type":      "application/json",
        },
        json={
            "model":       model or "claude-haiku-4-5-20251001",
            "system":      system,
            "messages":    msgs,
            "temperature": temperature,
            "max_tokens":  max_tokens,
        },
        timeout=120,
    )
    rsp.raise_for_status()
    return rsp.json()["content"][0]["text"]


if __name__ == "__main__":
    print(query_llm([
        {"role": "user", "content": "Say hi in three words."},
    ]))
\end{lstlisting}

\noindent
A first call against a local Ollama, with \fname{.env} set to
\code{LLM\_BACKEND=ollama} and \code{OLLAMA\_MODEL=gemma4:e4b}:

\begin{lstlisting}[language=bash]
$ python query_llm.py
Hi there, friend!
\end{lstlisting}

\noindent
Switching to a hosted backend is two lines in \fname{.env}:

\begin{lstlisting}[caption={Switching to OpenAI by editing \fname{.env}.},captionpos=b]
LLM_BACKEND=openai
OPENAI_API_KEY=sk-...
\end{lstlisting}

\noindent
\dots\ and re-running the same script:

\begin{lstlisting}[language=bash]
$ python query_llm.py
Hello, world there!
\end{lstlisting}

\noindent
Three things about the design.

First, the wrapper accepts and returns plain Python --- a list of
message dicts in, a string out. No SDK objects, no streaming
iterator types, no tool-call envelopes. Per
\S\ref{sec:intro-approach}, the book's chapters compose this
primitive into more interesting behavior rather than reach for
richer abstractions.

Second, the dispatcher is one dict and one lookup. Switching
backends is an environment-variable change, never a code change.
The reader who has set up only the local backend can ignore the
hosted branches; the reader who has set up only the hosted backend
can ignore the local branch. Neither has to be removed for the
other to work.

Third, the function signature does not include tool calls. Tool
calls add structural complexity (the response can be either a text
reply or a list of tool requests) and naming complexity (which
provider's tool-call schema do we adopt as the unified shape?).
Chapter~\ref{ch:toolcalls} introduces a richer wrapper that handles
tool calls; it builds on the same backend implementations defined
here.

\subsection{A one-shot convenience: \code{ask\_llm}}
\label{sec:work-wrapper-ask}

The full chat shape of \code{query\_llm} is the right tool when
the call has structure --- a non-trivial system prompt, several
turns of history, multiple roles in flight. Many of the book's
later examples are not like that. They are one-shot: a single
string in, a single string out. For those, building a one-element
messages list is busywork that obscures what the example is
actually doing.

We therefore ship a thin convenience wrapper alongside
\code{query\_llm} in the same file. \code{ask\_llm} takes a string
prompt (and, optionally, a system prompt) and returns the model's
reply directly.

\begin{lstlisting}[language=Python,caption={The convenience wrapper. Lives in \fname{manuscript/scripts/query\_llm.py} alongside \code{query\_llm} and shares the same backends.},label=lst:ask-llm]
def ask_llm(
    prompt: str,
    *,
    system_prompt: Optional[str] = None,
    model:         Optional[str] = None,
    temperature:   float = 0.0,
    max_tokens:    int = 2048,
) -> str:
    """One-shot string query. Convenience wrapper around `query_llm`."""
    messages: list[dict[str, str]] = []
    if system_prompt is not None:
        messages.append({"role": "system", "content": system_prompt})
    messages.append({"role": "user", "content": prompt})
    return query_llm(
        messages,
        model=model,
        temperature=temperature,
        max_tokens=max_tokens,
    )
\end{lstlisting}

\noindent
The two functions are interchangeable; the calls

\begin{lstlisting}[language=Python]
ask_llm("How many r's are in 'strawberry'?")

query_llm([{"role": "user", "content": "How many r's are in 'strawberry'?"}])
\end{lstlisting}

\noindent
produce identical requests. Through the rest of the book, where
an example needs only one user message we use \code{ask\_llm};
where the structure of the call matters --- multi-turn agents,
explicit system prompts, the tool-call loop of
Chapter~\ref{ch:toolcalls} --- we use \code{query\_llm} directly.

The reader who has run the smoke test of \S\ref{sec:work-deploy}
against at least one backend, and seen either \code{ask\_llm} or
\code{query\_llm} return their first answer, is ready for the
rest of the book.

% Chats: the Message Structure has been moved up, ahead of
% Interacting with LLMs, so the implementation sections can lean
% on the role vocabulary already in the reader's hands.

\section{Structured Outputs}
\label{sec:structured}

By default a language model emits a stream of tokens that decode
to free-form natural language. Many downstream uses need more: a
JSON object with a fixed shape, an XML document, a SQL query, a
function signature. Three families of techniques are in common
use, in increasing order of guarantees:

\begin{enumerate}[label=(\arabic*)]
\item \textbf{Prompting.} The system prompt instructs the model
  to emit, say, ``a single JSON object on one line, no prose, no
  code fences''. This works most of the time on a sufficiently
  large model and almost always with one or two few-shot
  examples. It is the cheapest option and the one most readers
  will start with.
\item \textbf{Constrained decoding.} The runtime intercepts the
  token stream and masks out any token that would violate a
  target grammar (\code{llama.cpp} grammars, \code{Outlines},
  \code{lm-format-enforcer}, the structured-output modes shipped
  by hosted providers). The output is \emph{guaranteed} to parse
  against the schema.
\item \textbf{Fine-tuning.} The model is trained on examples of
  the desired structured format. Used together with constrained
  decoding it produces the most reliable output, at the cost of
  a training run.
\end{enumerate}

\noindent
A tool call is the canonical application of structured outputs: a
JSON object with a known schema, intercepted by the runtime,
dispatched, and answered. Chapter~\ref{ch:toolcalls} treats the
tool call as the unit of work and assumes structured-output
reliability is a solved problem at the layer below.

\todo{Expand: a worked example of constrained decoding against a
local model; a comparison of the schema languages used by major
providers (OpenAI's \code{response\_format}, Anthropic's tool-use
schema, Gemma~4's auto-derived schema from Python type hints);
the trade-off between prompt-engineering reliability and
constrained-decoding rigidity.}

\section{Cost, Caching, and Latency}
\label{sec:work-cost}

\todo{The operational triple. Three subsections.}

\subsection{Cost}
\label{sec:work-cost-cost}

\todo{Per-token pricing. The asymmetry between input and output
tokens (input is much cheaper, often by a factor of three to
five). How to estimate a session cost from a prompt template and
expected output length. The trap of agentic loops: cost can scale
with the number of iterations, not with the user-perceived task
size. Forward-reference the context discussion in
Chapter~\ref{ch:toolcalls}.}

\subsection{Prompt and KV caching}
\label{sec:work-cost-cache}

\todo{Why caching matters disproportionately for long-running
agents (the same system prompt and tool catalog is sent every
turn). The provider-side surface: explicit cache APIs (Anthropic),
automatic prefix caching (OpenAI, several open-source servers),
nothing-at-all. How to design a system prompt to be
cache-friendly: stable prefix, dynamic suffix.}

\subsection{Latency}
\label{sec:work-cost-latency}

\todo{Time-to-first-token vs total-tokens-per-second; the two
matter for different use cases (interactive UX cares about TTFT,
batch agents care about total throughput). Streaming as a UX
pattern. The latency tax of reasoning models (the thinking
tokens are still tokens). Practical numbers will date; cite the
ratios rather than absolute figures.}
